\documentclass{ximera}
\graphicspath{{./content/hw_05_tnb_accleration/graphics/}{./graphics/}}
\title{Homework 5: Moving Frames and Acceleration}
\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

SOME PROBLEMS COPIED FROM HANDOUTS - NOT SURE OF THEIR SOURCE

\section*{Completion Packet}

\begin{problem}
%copied from handout
Find the velocity, acceleration and speed of a particle with the position function $\mathbf{x}(t) = (t, t^{2}, 2)$. Sketch the path of the particle \emph{by hand}, and draw the velocity and acceleration vectors on your graph for  $t=1$.
\end{problem}

\begin{problem}
%copied from handout
Find the velocity, acceleration and speed of a particle with the position function $\mathbf{x}(t) = (t, 2 \cos{t}, \sin{t})$. Sketch the path of the particle \emph{by hand}, and draw the velocity and acceleration vectors on your graph for  $t=0$.
\end{problem}

\begin{problem}
Find the moving frame for the path $\vec{x}(t) = (2t, 6\sin(3t), -6\cos(3t))$.
\end{problem}

\begin{problem}
Find the the acceleration of the path $\vec{x}(t) = (t,t^2)$, and express it as a linear combination of the unit tangent vector and the unit normal vector.
\end{problem}

\begin{problem}
Find the the acceleration of the path $\vec{x}(t) = (e^t, t)$, and express it as a linear combination of the unit tangent vector and the unit normal vector.
\end{problem}

\begin{problem}
Suppose we have a path $\vec{x}(t)$ and a point $\vec{c}$ in $\mathbb{R}^3$ such that 
$\|\vec{x}(t)-\vec{c}\|=R$ for some constant $R$ and for all $t$.
\begin{enumerate}
\item Geometrically, what can you say about this path?
\item Prove that $\vec{v}(t)$ is perpendicular to $\vec{x}(t)-\vec{c}$.
\item If $\vec{x}(t)$ has constant speed, show that $\kappa(t)$ is nonzero for all $t$.
\end{enumerate}
\end{problem}

\begin{problem}
%copied from handout
Prove that if a particle moves with constant speed, then the velocity and acceleration vectors are orthogonal.
\end{problem}

\begin{problem}
%copied from handout
Find the tangential and normal components of the acceleration vector for $\mathbf{x}(t)=(1+t,t^{2}-2t)$.
\end{problem}

\begin{problem}
%copied from handout
Find the tangential and normal components of the acceleration vector for $\mathbf{x}(t)=(t, t^{2}, 3t)$.
\end{problem}

\begin{problem}
%copied from handout
Compute the following limits. Show all work. (``Wolfram Alpha" is not work.)
\begin{enumerate}
\item $\displaystyle \lim_{x\to 3} \cfrac{x^2-9}{x^2+2x-3}$
\item $\displaystyle \lim_{x\to \pi^-} \ln(\sin{x})$
\item $\displaystyle \lim_{v\to 4^+} \frac{4-v}{|4-v|}$
\item $\displaystyle \lim_{x\to \infty} (\sqrt{x^2+4x+1} - x)$ 
\item $\displaystyle \lim_{x\to \infty} e^{x-x^2}$
\item $\displaystyle \lim_{x\to 0} \frac{\sin{2x}}{x}$ 
\end{enumerate}
\end{problem}

\section*{Graded Problems}

\begin{problem}
Find the moving frame for the path $\vec{x}(t) = (e^t\cos(t), e^t\sin(t), 5)$.
\end{problem}

\begin{problem}
Prove that 
\[\frac{|(\vec{v}\times\vec{a})\cdot \vec{a}'|}{\|\vec{v}\times\vec{a}\|^2} = \frac{\left\|d\vec{B}/dt\right\|}{|ds/dt|}.\]
\end{problem}

\section*{Professional Problem}

\begin{problem}
\begin{enumerate}
%copied from handout
\item Beginning with the formula $\mathbf{a}=s'' \mathbf{T} + (s')^{2}\kappa \mathbf{N}$, cross both sides with the velocity vector $\mathbf{v}=s'\mathbf{T}$ and derive the formula
\[ \kappa = \frac{||\mathbf{v} \times \mathbf{a}||}{(s')^{3}}. \]
Make sure to write your solution in your own words!

\item Let $y=f(x)$.  The graph of $y=f(x)$ is therefore a plane curve.  Assuming $f \in C^2$ (that is, it is twice differentiable and its second derivatives are continuous), use the formula from (a) to prove the following well-known formula for the curvature of such a curve:
\[ \kappa(x) = \dfrac{|f''(x)|}{ \left[ 1 + (f'(x))^2 \right]^{3/2}}. \]

\end{enumerate}
\end{problem}



\end{document}