\documentclass{ximera}
\graphicspath{{./content/hw_12_extrema/graphics/}{./graphics/}}
\title{Homework 12: Extrema}
\begin{document}
\begin{abstract}
\end{abstract}
\maketitle

\section*{Graded Problems}

\begin{problem}
 Consider the function $f(x,y) = (y^2-x)(2y^2-x)$. \begin{enumerate}
\item Show that $f(x,y)$ has a critical point at the origin.
\item Show that $f(x,y)$ has a local minimum along any line through the origin. That is, show that for constant $(a,b)\neq (0,0)$, that the function $g(t)=f(at,bt)$ has a local minimum at $t=0$.
\item Show that $f(x,y)$ does not have a local minimum at the origin.
\end{enumerate}
\end{problem}

\begin{problem}
\begin{enumerate}
%For next year: consider restricting to n=3, providing scaffolding to help with substitution.
\item Let $x_1,x_2,...,x_n$ be nonnegative numbers such that their sum is constant. That is, $x_1+x_2+\cdots +x_n=C$ for some constant $C$. Show that the product $x_1x_2\cdots x_n$ is a maximum if and only if $x_1=x_2=\cdots =x_n = C/n$.
\item Using the result of part (a), show that if $x_1,x_2,...,x_n$ are nonnegative numbers such that $x_1+x_2+\cdots +x_n=1$, then $x_1x_2\cdots x_n\leq 1/n^n$.
\end{enumerate}
\end{problem}

\section*{Professional Problem}

\begin{problem} Final draft of your project, worth two professional problems.
\end{problem}

\section*{Completion Packet}

\begin{problem}
Find and classify all critical points of each function. When using the Hessian fails to classify points, find another method.

\begin{enumerate}

\item $f(x,y) = x^3 + y^3 + 12xy$

\item $f(x,y) = x^2+y^2+\frac{1}{x^2y^2}$, for $xy\neq 0$

\item $f(x,y) = (x+y)^2+x^4$

\item $f(x,y) = (x+y)e^{-xy}$

\item $f(x,y,z) = 2x^2+y^2+z^2-xz+xy$

\item $f(x,y,z) = xy+xz$
\end{enumerate}

\end{problem}

\begin{problem} What are the conditions on $a,b,c$ for $f(x,y) = ax^2+bxy+cy^2$ to have a...\begin{enumerate}
\item ...local minimum at the origin?
\item ...local maximum at the origin?
\item ...saddle point at the origin?
\end{enumerate}
\end{problem}

\begin{problem} For nonzero constants $a$ and $b$, consider the function $f(x,y) = ax^{-1}+by^{-1}+xy$.\begin{enumerate}
\item Find the (single) critical point of this function.
\item What are the conditions on $a$ and $b$ for the critical point to be a local minimum? A local maximum? A saddle point?
\end{enumerate}
\end{problem}

\begin{problem} %For next year: consider providing scaffolding to help with substitution.
Find the shortest distance between a point on the surface $(x-2)^2+(y-3)^2+z^2=1$ and the origin in $\mathbb{R}^3$.
\end{problem}

\begin{problem} \begin{enumerate}
%For next year: consider restricting to n=3, providing scaffolding to help with substitution.
\item Let $x_1,x_2,...,x_n$ be positive numbers such that their product is constant. That is, $x_1x_2\cdots x_n=C$ for some constant $C$. Show that the sum $x_1+x_2+\cdots+x_n$ is a \emph{local} minimum if and only if $x_1=x_2=\cdots =x_n$.
\item The local minimum which you found in part (a) is, in fact, an absolute minimum (you do not need to show this). Using this fact, show that if $x_1,x_2,...,x_n$ are positive numbers such that $x_1x_2\cdots x_n=1$, then $x_1+x_2+\cdots+x_n\geq n$.
\end{enumerate}
\end{problem}



\end{document}
